---
title: "Practical Machine Learning - Project Course"
author: "Arturo Equihua"
date: "June 11, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Synopsis
Poner aqui el resumen de lo que haya visto.

##Introduction
This document describes a proposed model that was built to predict the "quality of workout done" (well done or done with mistakes), based on several variables that were measured on subjects wearing special devices (Nike Fuelband and so on), while performing specific weightlifting movements.

The source data comes from the following study, which also has additional information about the nature of it: 

*Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human) . Stuttgart, Germany: ACM SIGCHI, 2013.*

##Dataset Structure and Preparation
The input data comes in CSV format, in two files (training set and testing set). The variable that we will predict is the "classe" variable (Class A means "well done", and classes B, C, D and E represent specific types of mistakes).

> More information about the dataset is available in <http://groupware.les.inf.puc-rio.br/har>, in the section "Weight Lifting Exercise Dataset".

The code to load the training and testing data is shown below:
```{r load_data, echo=TRUE, results='hide', warning=FALSE, cache=FALSE}
  library(caret)
  wltrain <- read.csv("pml-training.csv")
  wlptest <- read.csv("pml-testing.csv")
```
In order to select the features for the prediction model (the entire set has 160 variables that could be used), a quick look at the training dataset suggests there are variables that could be removed as they have many NA values in them. We can also eliminate variables that are not relevant for the problem we are trying to solve, or have little variance (repeat the same value too much). We then do the variable cleansing as shown below:

```{r load_data, echo=TRUE, results='hide', warning=FALSE, cache=FALSE}

```

The raw data needs then to be filtered for the relevant columns, and also some treatment is required on the date, people fatality/injury and economic damage figures. In particular, the date needs to be converted from character to date, a summed column needs to be created to combine deceases and injuries, and the economic figures need to be standardized depending on the unit (K for thousands, M for millions or B for billions). The following code shows the transformations done:
```{r select_data, echo=TRUE, results='hide', warning=FALSE, message=FALSE,cache=FALSE}
  library(dplyr)
  library(stringr)
  library(lubridate)

  # Select the columns
  stormdata <- select(rawdata,BGN_DATE:EVTYPE,FATALITIES:CROPDMGEXP)

# Convert the date data to a date column (no need for time data)
  stormdata <- mutate(stormdata,BGN_DATE=word(BGN_DATE,1))
  stormdata <- mutate(stormdata,BeginDateTime=mdy(BGN_DATE)) 

# Add a Year column
  stormdata <- mutate(stormdata,Year=year(BeginDateTime))

# Convert the US dollar amounts based on the EXP character field (K, M or B to denote thousands, millions or billions)
  convert <- function(num, code) {
    newnum = num
    if (code=="K")
       newnum=num*1000
    else if (code=="M")
       newnum=num*1000000
    else if (code=="B")
       newnum=num*1000000000
    newnum
  }
  stormdata <- mutate(stormdata, 
                      PropDamage=mapply(convert,PROPDMG,PROPDMGEXP),
                      CropDamage=mapply(convert,CROPDMG,CROPDMGEXP),
                      SumDamage=PropDamage+CropDamage)

# Add a column that combines fatality and injury figures 
  stormdata <- mutate(stormdata,
                      AffectedPeople=FATALITIES+INJURIES)

# Take out columns that are no longer needed
  stormdata = select(stormdata,-(BGN_DATE:TIME_ZONE))
  stormdata = select(stormdata,-(PROPDMG:CROPDMGEXP))
```

The steps above produced a filtered set of data. For the purposes of the analysis, which considers information at the country level, the data needs to be summarized at the Event Type Level.

This is the code to produce the summarized data:

```{r summarize_data, echo=TRUE, results='hide', warning=FALSE, message=FALSE,cache=FALSE}

# Summarize by Year and Event Type

  library(plyr)

  # Group the Fatalities, Injuries, PropDamage, CropDamage, SumDamage and Affected People figures
  sumed=ddply(stormdata,c("EVTYPE"),
              summarise,
              Quantity=length(EVTYPE),
              Fatalities=sum(FATALITIES),
              Injuries=sum(INJURIES),
              AffectedPeople=sum(AffectedPeople),
              PropDamage=sum(PropDamage),
              CropDamage=sum(CropDamage),
              SumDamage=sum(SumDamage))
```

The following is the resulting list of data elements for the rest of the analysis:

- EVTYPE (chr) : Short name of the type of event (e.g. Tornado, Hurricane)
- Quantity (num)  : Number of events per event type across the period
- FATALITIES (num) : Number of deceased people as a result of the event
- INJURIES (num) : Number of injured people as a result of the event
- AffectedPeople (num) : Combined fatality and injury figures
- PropDamage(num) : Amount of property damage in USD
- CropDamage(num) : Amount of crop damage in USD
- SumDamage(num) : Total amount of damage in USD

The following section presents the findings of the analysis of this set of data. These outcomes constitute a basis for initial discussions about that is the priority that needs to be set for the attention of the most severe weather events at the country level.

##Results Obtained
### Impact in People Fatality or Injury
By aggregating the historic information of the 1950-2011 period in terms of injured and deceased people, it is clear that **tornadoes** are, by far, the most destructive events for population:

```{r plot_health, echo=TRUE, results='hide', warning=FALSE, message=FALSE,cache=FALSE,fig.cap="Top 8 Weather Events in terms of people health"}

# Reorder by people impact
  library(dplyr)
  sample=head(arrange(sumed, desc(AffectedPeople)),8)

# For ggplot, the X axis needs to be explicily reordered to show the bars
# in order of size
  sample$EVTYPE = with(sample,reorder(EVTYPE,AffectedPeople))

  library(reshape2)
  melted=melt(sample,id=c("EVTYPE"),measure.vars=c("Fatalities","Injuries"))
  colnames(melted)=c("EvType","Impact","AffectedPeople")

# Produce the plot for top 8 events
# Did two barplots in the same grid: One with the total and another with the 
# stacked bar
  library(ggplot2)
  ggplot(data=sample, aes(x=EVTYPE, y=AffectedPeople)) +
        geom_bar(colour="black", fill="#DD8888", width=.8,stat="identity")+
        ggtitle("People Highest-impact Types of Weather Events (1950-2011)") +
        ylab("Affected People(Fatality + Injured)") +
        xlab("Type of Event") +
        scale_x_discrete(limits=rev(levels(sample$EVTYPE))) +
        geom_bar(aes(x=melted$EvType,y=melted$AffectedPeople,
                     fill=melted$Impact),stat="identity",width=.5)+
        scale_fill_manual(values=c("dodgerblue4", "darkolivegreen4",
                              "darkorchid3", "goldenrod1")) +
        theme(panel.background = element_rect(fill = 'grey75'),
              plot.background=element_rect(fill="darkseagreen"),             
              panel.grid.major = element_line(colour = "orange", size=1),
              panel.grid.minor = element_line(colour = "blue"),
              legend.title=element_blank(),
              axis.title.x = element_text(face="bold"),
              axis.title.y = element_text(face="bold",size=10),
              axis.text.x = element_text(angle=90, vjust=1,colour="blue"),
              axis.text.y = element_text(angle=90, vjust=1,colour="blue"))
  
```

Other events that unfortunately lead to the most fatalities and injured people are, in that order, **Excessive Heat, Thunderstorms with Wind, Floods, Lightning storms, Heat, Flash floods and Ice Storms**. These are the events for which authorities should prepare the most in terms of public health planning.

## Economic Impact
When looking at weather events in terms of economic damage (destroyed private property, loss of farm crops), the findings vary significantly:

```{r plot_economic, echo=TRUE, results='hide', warning=FALSE, message=FALSE,cache=FALSE,fig.cap="Top 8 Weather Events in terms of economic damage"}

# Reorder by people impact
  library(dplyr)
  sample=head(arrange(sumed, desc(SumDamage)),8)
  sample=mutate(sample,PropDamage=PropDamage/1e+6,
                CropDamage=CropDamage/1e+6,
                SumDamage=SumDamage/1e+6)

# For ggplot, the X axis needs to be explicily reordered to show the bars
# in order of size
  sample$EVTYPE = with(sample,reorder(EVTYPE,SumDamage))

  library(reshape2)
  melted=melt(sample,id=c("EVTYPE"),measure.vars=c("PropDamage","CropDamage"))
  colnames(melted)=c("EvType","Impact","EconomicDamage")

# Produce the plot for top 8 events
# Did two barplots in the same grid: One with the total and another with the 
# stacked bar
  library(ggplot2)
  ggplot(data=sample, aes(x=EVTYPE, y=SumDamage)) +
        geom_bar(colour="black", fill="#DD8888", width=.8,stat="identity")+
        ggtitle("Economic Highest-impact Types of Weather Events (1950-2011)") +
        ylab("Economic Damage (million USD)") +
        xlab("Type of Event") +
        scale_x_discrete(limits=rev(levels(sample$EVTYPE))) +
        geom_bar(aes(x=melted$EvType,y=melted$EconomicDamage,
                     fill=melted$Impact),stat="identity",width=.5)+
        scale_fill_manual(values=c("dodgerblue4", "darkolivegreen4",
                              "darkorchid3", "goldenrod1"),
                          labels=c("Property Damage","Crop Damage")) +
        theme(panel.background = element_rect(fill = 'grey75'),
              plot.background=element_rect(fill="lightblue"),             
              panel.grid.major = element_line(colour = "orange", size=1),
              panel.grid.minor = element_line(colour = "blue"),
              legend.title=element_blank(),
              axis.title.x = element_text(face="bold"),
              axis.title.y = element_text(face="bold",size=10),
              axis.text.x = element_text(angle=90, vjust=1,colour="blue"),
              axis.text.y = element_text(angle=90, vjust=1,colour="blue"))
  
```


What the above suggests is that water-related events such as **floods and hurricanes** are the most damaging events for the economy of the country.  In particular, floods alone have caused economic damages for over **150 billion USD** in the 60-year period under analysis. It also can be seen that the damage to property (buildings, infrastructure) is the highest in most of the cases. On the other hand, drought is the most damaging weather condition for farmers.
